Fast Sorting on Modern Computers

### 2005-08-30 docs/tech_report/conclusions.tex
formatting fixes. -- and "\.  "
spells checks

Sedgewick proposed removing individual insertion sorts in quicksort with one
elegant sweep at the end. LaMarca proposed replacing the large number of
insertion sorts in order to reduce cache misses. We note that attempting to move
the pivots, as Sedgewick does, actually increases both the number of branches
and the rate of mispredictions. We conclude that LaMarca's improvment, which
slightly improves quicksorts cache performance, also slightly improves it's
branch prediction performance.

@@

We also analysed the effects of different medians on quicksort. We deterine that
not using a median to choose a pivot makes the branches more predictable than
they are biased, though at a cost of a large increase in instructions (and also
significantly greater possability of a worst-case $O(N^2)$ sort). We also saw
that the more keys considered when determining the median, the lower the number
of branches. However, this can result in an increase in the rate of branch
mispredictions. We determine that reducing entropy is not the best way of making
quicksort perform faster, and that using a median-of-three actually achieves the
higher performance.

@@

We are able to accurately predict the number of branch mispredictions in
shellsort, using our results from insertion sort. We also found that the
improvement mode to shellsort, seperating the final loop from the rest, results
in keys being moved 1.07 places, up from 1.03, contributing to its improved
performance.

>>>>

Sedgewick proposed removing individual insertion sorts in quicksort with one
elegant sweep at the end. LaMarca proposed replacing the large number of
insertion sorts in order to reduce cache misses. We note that attempting to move
the pivots, as Sedgewick does, actually increases both the number of branches
and the rate of mispredictions. We conclude that LaMarca's improvement, which
slightly improves quicksorts cache performance, also slightly improves it's
branch prediction performance.

@@

We also analysed the effects of different medians on quicksort. We determine that
not using a median to choose a pivot makes the branches more predictable than
they are biased, though at a cost of a large increase in instructions (and also
significantly greater possibility of a worst-case $O(N^2)$ sort). We also saw
that the more keys considered when determining the median, the lower the number
of branches. However, this can result in an increase in the rate of branch
mispredictions. We determine that reducing entropy is not the best way of making
quicksort perform faster, and that using a median-of-three actually achieves the
higher performance.

@@

We are able to accurately predict the number of branch mispredictions in
shellsort, using our results from insertion sort. We also found that the
improvement mode to shellsort, separating the final loop from the rest, results
in keys being moved 1.07 places, up from 1.03, contributing to its improved
performance.

### 2005-08-30 docs/tech_report/elementary.tex
spells checks

The level 1 cache should have bad performance as a result. Since the level 1
cache is smaller that the arrays tested, it should perform analagiously to a
level 2 cache being tested with a much larger array.

@@

In Figure \ref{Predictability of branches in bubblesort compared to sweep
number} we see that on the early iterations the number of correct predictions is
very close to the number of executed comparison branches. The comparison branch
is highly predictable because bubblesort's search for the smallest element is
similar to selection sort. However, with successive sweeps even the unsorted
part of the array becomes increasingly sorted and the branch becomes more
unpredictable. We see that the number of executed comparison branches falls
linearly, but that the number of correctly predicted comparison branches falls
much more quickly. Eventually, the array becomes so close to fully sorted that
the branch becomes increasingly predictable in the other direction, and the
number of correct predictions again approaches the number of executed branches.
However, during the intermediate stage, a huge number of branch mispredictions
have already occured. Paradoxically, by partially sorting the array, bubblesort
greatly increases the number of branch mispredictions and thus greatly increases
its running time. Bubblesort would be substantially faster if, instead of
attempting partially sort these intermediate elements, it simply left them
alone.

>>>>

The level 1 cache should have bad performance as a result. Since the level 1
cache is smaller that the arrays tested, it should perform analogously to a
level 2 cache being tested with a much larger array.

@@

In Figure \ref{Predictability of branches in bubblesort compared to sweep
number} we see that on the early iterations the number of correct predictions is
very close to the number of executed comparison branches. The comparison branch
is highly predictable because bubblesort's search for the smallest element is
similar to selection sort. However, with successive sweeps even the unsorted
part of the array becomes increasingly sorted and the branch becomes more
unpredictable. We see that the number of executed comparison branches falls
linearly, but that the number of correctly predicted comparison branches falls
much more quickly. Eventually, the array becomes so close to fully sorted that
the branch becomes increasingly predictable in the other direction, and the
number of correct predictions again approaches the number of executed branches.
However, during the intermediate stage, a huge number of branch mispredictions
have already occurred. Paradoxically, by partially sorting the array, bubblesort
greatly increases the number of branch mispredictions and thus greatly increases
its running time. Bubblesort would be substantially faster if, instead of
attempting partially sort these intermediate elements, it simply left them
alone.

### 2005-08-30 docs/tech_report/merge.tex
spells checks

Double aligned multi-mergesort is an even more complicated algorithm. It sorts
the arrays fully in the cache, leaving them bitonically arranged (sorted into
sub-arrays, sorted into alternatey increasing and decreasing order). The final
merge is performed separately, and hence no bias for the forward branches, as
there was in tiled mergesort. The final sort is performed with a $k$-way merge,
which is implemented using a heap data structure. This is shown in Figures
\ref{mergesorts-5} and \ref{mergesorts-6}. The `\n{Comparison}' branches are
comparisons from the merge. The labels in these graphs are the same as in Figure
\ref{mergesorts-4} and \ref{base and memory-tuned heapsorts-2}.

>>>>

Double aligned multi-mergesort is an even more complicated algorithm. It sorts
the arrays fully in the cache, leaving them bitonically arranged (sorted into
sub-arrays, sorted into alternately increasing and decreasing order). The final
merge is performed separately, and hence no bias for the forward branches, as
there was in tiled mergesort. The final sort is performed with a $k$-way merge,
which is implemented using a heap data structure. This is shown in Figures
\ref{mergesorts-5} and \ref{mergesorts-6}. The `\n{Comparison}' branches are
comparisons from the merge. The labels in these graphs are the same as in Figure
\ref{mergesorts-4} and \ref{base and memory-tuned heapsorts-2}.

### 2005-08-30 docs/tech_report/quick.tex
spells checks

LaMarca notes that quicksort is already very cache efficient\footnote{We
speculate wildly at this point and suggest that since an average of 25\% of
computing time is spent sorting (see \cite{TODO}), the development of caches has
been subconsicously affected by the cache performance of quicksort.}, as it has
excellent spatial and temporal locality. Once it no longer fits in the cache,
however, the number of misses should increase significantly. This is especially
true in the case of the base quicksort, which does an insertion sort across then
entire array at the end.

@@

In order to investigate this, we implemented a simple version of quicksort that
uses the middle element as the pivot rather than a median-of-three\footnote{It
is important to note that selecting the pivot based on only a single element can
be particularly inefficient because it removes the natural sentinels at the top
and bottom of the regions to be partitioned that we get from median-of-three. In
Figure \ref{quick-predictors2} we see that this results in an additional branch
to handle the edgecase without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort
are random, so the value of this element will also be random. Figure \ref{base
quicksort without a median, with median of 3, and with pseudo-medians of 5, 7
and 9-1} shows the behaviour of each of the important branches in this code. We
see that the \n{i} and \n{j} branches are indeed taken about twice as often as
not-taken, and the \n{i} and \n{j} branches are correctly predicted
approximately 71\% of the time.

>>>>

LaMarca notes that quicksort is already very cache efficient\footnote{We
speculate wildly at this point and suggest that since an average of 25\% of
computing time is spent sorting (see \cite{TODO}), the development of caches has
been subconsciously affected by the cache performance of quicksort.}, as it has
excellent spatial and temporal locality. Once it no longer fits in the cache,
however, the number of misses should increase significantly. This is especially
true in the case of the base quicksort, which does an insertion sort across then
entire array at the end.

@@

In order to investigate this, we implemented a simple version of quicksort that
uses the middle element as the pivot rather than a median-of-three\footnote{It
is important to note that selecting the pivot based on only a single element can
be particularly inefficient because it removes the natural sentinels at the top
and bottom of the regions to be partitioned that we get from median-of-three. In
Figure \ref{quick-predictors2} we see that this results in an additional branch
to handle the edge-case without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort
are random, so the value of this element will also be random. Figure \ref{base
quicksort without a median, with median of 3, and with pseudo-medians of 5, 7
and 9-1} shows the behaviour of each of the important branches in this code. We
see that the \n{i} and \n{j} branches are indeed taken about twice as often as
not-taken, and the \n{i} and \n{j} branches are correctly predicted
approximately 71\% of the time.
